---
title: "Introduction to Coding in R + Git"
author: "Sean Li and Ben Thorpe"
date: "1/25/2022"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
Note: this code chunk is just default setup stuff RStudio does. Don't worry about it (you can keep it)


If you have not downloaded R, git and RStudio, please refer to the README.md file in the repository and complete those steps before proceeding. You will only need to do this once. 

## Packages

Here we load in packages, which is basically a bundle of pre-made code that helps users create things. For example, the tidyverse package contains built-in functions that allow for data visualization and data wrangling. The Lahman package contains baseball data we are working with. 

```{r}
library(tidyverse) #package for data wrangling + data viz
library(Lahman) #package with data
```

## Essentials

Code chunks are the sections in gray surrounded by the triple hash marks and r in brackets. They contain code that is run if you click the green arrow on the top right of the code chunk. 

You can insert more code chunks by pressing the green button Insert or pressing option + Command + I. 

Naming code chunks is when you have something next to the r in the brackets of your code chunk. This is essential for documentation. 

Knitting is the process of putting together your R Markdown Document into a format that can be exported. Currently, we have it set to HTML and PDF. You set the format in the output section up in the header of the Rmd. 

Commenting your code can be done by using a #, you will see the text turn green after. This must be done in the code chunk. Text after the # on that line will not be run when you compile. 

## Introduction (a quick primer on R and RStudio)

### Vectors

A vector is the basic building block in R. 

the c() stands for concatenate. 

```{r vectors}
c(TRUE, FALSE, TRUE) #logical
c(4, 1, 0) #integer
c(pi, 5 / 2, 6 * 4, -3) #double/numeric
c("intro", "to", "r") #character
```

The four main vector types you will work with are `logical`, `integer`,
`double`, and `character`. Each element in a vector should be of the same type or
we can run into trouble. R will automatically coerce (force) a type into another type if 
conflict arises. 

Some vector properties:

1. Each element of a vector must be the same type, check the type with 
   `typeof()`

2. Vector indexing begins at 1, not 0

3. Vectors can be subset by position

### Basic Calculations 

Here are some basic functions you can apply to a vector.

mean(x) - mean
median(x) - median
sd(x) - standard deviation
var(x) - variance
cor(x, y) - correlation 
cov(x, y) - covariance


```{r indexing}
v <- c(3,5,6,0,9)


v[2]
mean(v)
summary(v)
v[1:4] 
v[c(1,4)]
v[v>5]
```

```{r math}
w <- c(4,23,5,1,3)

v+w
v*2 + mean(w)
(w - mean(w)) / sd(w) # z scores
```

Adding elements

```{r adding elements}
names <- c("sean", "ben")
names <- c(names, "jack")
```



### Data frames (tibbles)

A data frame (or tibble) is a rectangular structure in R where each column is
a vector. It is a very common structure to work with in doing data wrangling
and visualization.

`mtcars` is the name of the dataframe/tibble. Simply writing it in a code chunk will display it. 
```{r cars}
mtcars
```

A tibble is a nicer form of a data frame, so generally we will coerce data
frames to tibbles with `as_tibble()`. This is what you usually have to do given raw sports data. 


### Object creation

Objects can be created with `<-` or `=`. However, using the assignment operator,
`<-` is considered best practice.

```{r objects}
a <- 10 ^ 4
b <- sqrt(pi)

total <- sum(c(4, 5, 0, 10))
coin_tosses <- sample(c("H", "T"), size = 12, replace = TRUE)
```

```{r display-objects}
a
b
total
coin_tosses
```

### Casting

Sometimes we need to change the typing of a variable. Remember there are 4 main data types: 
integer, character, numeric, logical. 

For example, often a number is cast as a character originally

```{r}
bad <- c("3","2", "8")

numbers <- as.numeric(bad)
```

There are other ones such as as.factor() and as.character().

### Try it yourself 1 

1. Create a vector with the names of your family members called 'family'. 

2. Create a vector with the ages of your family members called 'family_ages'

3. What is the average age of your family members?

4. Index to the 2nd to last person in your family names vector. Who is it?

## Data Frames

Data frames are essentially objects in R to store tabular data, like an excel sheet.

We can construct a dataframe from vectors. 

```{r creating-df}

df = data.frame(names,numbers) #columns of vectors go into this function
view(df)
```


We'll work with some baseball data from the `Lahmen` package.

Notice how you can see each column header and the <type>. 
```{r display-table}
Teams
```

```{r tibble}
teams <- as_tibble(Teams)
```

Take a `glimpse()` at `teams`. 

```{r glimpse}
glimpse(teams)
```


### NA Values

Lets look at missing values and how to detect them

```{r na-vals}
anyNA(teams$yearID)
filtered <- na.omit(teams) #removes rows with any NA values in any columns
```


To do further manipulations we need to introduce the pipe operator - `%>%`.
This operator takes the object on the left-hand-side and passes it into
the object on the right-hand-side. Rather than write `f(x, y)` one could
write `x %>% f(y)`.


## Summary Statistics

Lets now go over ways to get some common summary statistics for dataframes. 

Some essential things you need to understand about each dataframe:
- What does each row represent?
- What does each column represent?
- Can we find the center and spread of a column? 

Number of rows, number of columns

```{r rows and columns}
nrow(mtcars) #mtcars is a built in dataset
ncol(mtcars)
length(mtcars)
```

Lets try to see the mean + median mpg of the cars in the dataset

We need to use the `$` key to denote which column. 
Applying a function to a column always take the form of:
function(dataframe$column)

```{r meanmed}
mean(mtcars$mpg)
median(mtcars$mpg)
sd(mtcars$mpg)
```

How many unique values of cylinder of cars in the dataset? 

```{r unique}
unique(mtcars$cyl)
```


The summary function gives a quick overview containing the min,max,median,mean,IQR of each column.
```{r all-sum}
summary(mtcars)
```


### Try it yourself 2

Using the dataframe teams loaded in already from earlier, xomplete the following exercises.

4. How many rows are in the dataframe? 

5. How many years does this dataset cover?

6. What is the highest value in the column `attendance` ? What is the type of the column?




## Simple Data Wrangling

If we have time - sneak peek into some easy data wrangling. 

### Select

Let's `select()` the runs, runs allowed, wins, year, and team name.

```{r}
teams %>% 
  select(R, RA, W, yearID, name)
```

`select()` allows you to choose variables (columns). Separate your choices with 
a comma.

### Filter

Let's `filter()` `teams` for those years since 1970.

```{r}
teams %>% 
  filter(yearID >= 1970)
```

`filter()` chooses rows based on conditions of variables.

### Arrange

Let's `arrange()` `teams` in descending order by wins.

```{r}
teams %>% 
  arrange(desc(W))
```

The default sorting with `arrange()` is ascending.

### Mutate

Let's create a new variable in `teams` called win percentage.

```{r}
teams %>% 
  mutate(win_pct = W / G)
```

`mutate()` allows you to create new variables in your data frame.

### Example

We can combine many of these steps together. Consider the example below.

```{r}
teams %>% 
  select(yearID, name, W, G) %>% 
  mutate(win_pct = W / G) %>% 
  arrange(desc(win_pct)) %>% 
  filter(G == 162)
```

## Data visualization

The `ggplot2` package is a package in `tidyverse` that is great for data
wrangling data frame objects. 

### Basics

Every `ggplot` object starts with a call to function `ggplot()`.

```{r}
ggplot()
```

Add some data and variables you want to map to aesthetics.

```{r}
ggplot(data = teams, aes(x = R, y = W))
```

Add a geometry. In this case we'll use `geom_point()` to create a 
scatterplot.

```{r}
ggplot(data = teams, aes(x = R, y = W)) +
  geom_point()
```

Finally, add some labels and a theme.

```{r}
ggplot(data = teams, aes(x = R, y = W)) +
  geom_point() +
  labs(x = "Runs scored", y = "Wins", title = "More runs means more wins!") +
  theme_minimal()
```

With `ggplot`, the plus operator, `+`, has a special meaning. It links together
the `ggplot2` functions to build your visualization.

## Pythagorean wins

Let's put together everything we know and create a scatterplot showing the
Pythagorean wins versus a team's win percentage for all teams since 2000.

```{r}
teams %>% 
  filter(yearID >= 2000) %>% 
  ggplot(aes(x = R ^ 2 / (R ^ 2 + RA ^ 2), y = W / G)) +
  geom_point(color = "blue") +
  geom_abline(intercept = 0, color = "red") +
  labs(x = "Pythagorean Wins", y = "Win percentage",
       title = "Teams above the line exceeded their Pythagorean wins") +
  theme_minimal(base_size = 14)
```

